Analyzing the influence of Selection on Genetic Programming's Generalization ability in Symbolic Regression

### 2022-06-26 docs/rmd/presentation/slides.Rmd
bugged latex fragment

---
title: "Analyzing the influence of Selection on Genetic Programming's Generalization ability in Symbolic Regression"
subtitle: "A comparison of epsilon-lexicase Selection and Tournament Selection"
author: "Roman Hoehn, B.Sc. Wirtschaftspaedagogik"
date: "2022-06-29"
output:
  beamer_presentation:
    colortheme: "beaver"
    fonttheme: "structurebold"
    #extra_dependencies: ["float"]
    keep_tex: true
    fig_caption: yes
    number_sections: true
    slide_level: 2
    #includes:
      #in_header: ./settings.tex
toc: yes
# toc_depth: 3
bibliography: ../bib/bib/ref.bib
csl: ../bib/csl/harvard1.csl
---

```{r setup, include=FALSE}
library(kableExtra)
library(magrittr)
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(fig.pos = "!H")
```

@@

knitr::kable(
  read.csv("../tables/csv/variables_energy_heating.csv"),
  digits=3,
  caption="Overview - Energy Heating data set"
)    %>% row_spec(0,bold=TRUE ) %>% kableExtra::kable_styling(latex_options = "hold_position")
```

@@

knitr::kable(
  read.csv("../tables/csv/gp_parameters.csv"),
  digits=3,
  caption="Evolutionary Parameters"
) %>% row_spec(0,bold=TRUE ) %>% kableExtra::kable_styling(latex_options = "hold_position")
```

@@

## Examples

@@

### Hypothesis testing

#### TODO: reformulate hypothesis
1. The usage of $\epsilon$-lexicase selection will result in models that perform significantly different than models that are evolved using tournament selection:
2. Statistical significant differences between the mean errors of training and testing data exist for both selection operators
3. No differences in program size exist between the distribution underlying the samples produced by $\epsilon$-lexicase and the distribution underlying samples of tournament selection

@@

## Descriptive Statistics

![Distribution of Errors](../plots/mean_error_boxplot_all.png)

>>>>

---
title: "Analyzing the influence of Selection on Genetic Programming's Generalization ability in Symbolic Regression"
subtitle: "A comparison of epsilon-lexicase Selection and Tournament Selection"
author: "Roman Hoehn, B.Sc. Wirtschaftspaedagogik"
date: "2022-06-29"
output:
  beamer_presentation:
    colortheme: "beaver"
    fonttheme: "structurebold"
    keep_tex: true
    fig_caption: yes
    number_sections: true
    slide_level: 2
toc: yes
bibliography: ../bib/bib/ref.bib
csl: ../bib/csl/harvard1.csl
---

```{r setup, include=FALSE}
library(kableExtra)
library(magrittr)
knitr::opts_chunk$set(echo = TRUE)
```

@@

knitr::kable(
  read.csv("../tables/csv/variables_energy_heating.csv"),
  digits=3,
  caption="Overview - Energy Heating data set"
) 
```

@@

knitr::kable(
  read.csv("../tables/csv/gp_parameters.csv"),
  digits=3,
  caption="Evolutionary Parameters"
)
```

@@

## Example

@@

* Does the usage of $\epsilon$-lexicase parent selection influence the generalization behavior of genetic programming in symbolic regression if compared to tournament selection?

### 3 Questions

1. Differences in average fitness for both algorithms?
2. Differences in fitness for training and testing data?
3. Differences in program size?

@@

## Distribution of Fitness

![](../plots/mean_error_boxplot_all.png)

## Finding 1

* The differences in average fitness of the final solutions between tournament selection and $\epsilon$-lexicase selection are highly statistical significant ($\alpha=0.01$)
* Tournament selection-based GP achieves a higher fitness for both training and testing data
* Unexpected results based on the reviewed literature [@epsilon_lexicase_main], [@https://doi.org/10.48550/arxiv.1709.05394]

## Evolution of Fitness

![](../plots/mean_error_combined.png)

## Finding 2

* The gap between training and testing error is not statistically significant for both selection algorithms
* Both algorithms achieve a slightly better performance for the training data
* No proof for overfitting

## Statistical Test

```{r, echo=FALSE, warning=FALSE}

df <- read.csv("../tables/csv/mwu_matrix_error.csv")

knitr::kable(
  df,
  format = "latex",
  booktabs = TRUE,
  digits=3,
  caption="Mean Error - P-Values (MWU)"
)

```

## Evolution of Size

\begin{figure}
\centering
\includegraphics{../plots/size_subplotted.png}
\end{figure}

## Finding 3

### 2022-06-26 docs/rmd/presentation/slides.tex
bugged latex fragment

\begin{table}[!h]

\caption{\label{tab:unnamed-chunk-1}Overview - Energy Heating data set}
\centering
\begin{tabular}[t]{l|l}
\hline
\textbf{Variable} & \textbf{Description}\\
\hline
X1 & Relative Compactness\\
\hline
X2 & Surface Area\\
\hline
X3 & Wall Area\\
\hline
X4 & Roof Area\\
\hline
X5 & Overall Height\\
\hline
X6 & Orientation\\
\hline
X7 & Glazing Area\\
\hline
X8 & Glazing Area Distribution\\
\hline
y1 & Heating Load\\
\hline
y2 & Cooling Load\\
\hline
\end{tabular}
\end{table}
\end{frame}

@@

\begin{frame}{Evolutionary Parameters}
\protect\hypertarget{evolutionary-parameters}{}
\begin{table}[!h]

\caption{\label{tab:unnamed-chunk-2}Evolutionary Parameters}
\centering
\begin{tabular}[t]{l|l}
\hline
\textbf{Parameter} & \textbf{Value}\\
\hline
Population Size & 500\\
\hline
Number of Generations & 100\\
\hline
Mutation Rate & 20\%\\
\hline
Crossover Rate & 80\%\\
\hline
Tournament Size & 3\\
\hline
Epsilon selection & automatic\\
\hline
Elite Size & 0\\
\hline
\end{tabular}
\end{table}
\end{frame}

@@

\begin{frame}{Function Set}
\protect\hypertarget{function-set}{}
\begin{table}

\caption{\label{tab:unnamed-chunk-4}Functions}
\centering
\begin{tabular}[t]{l|r}
\hline
Function & Arity\\
\hline
Addition & 2\\
\hline
Subtraction & 2\\
\hline
Multiplication & 2\\
\hline
Negation & 1\\
\hline
Sine & 1\\
\hline
Cosine & 1\\
\hline
Protected Division & 2\\
\hline
Protected Natural Logarithm & 1\\
\hline
Protected Square Root & 1\\
\hline
\end{tabular}
\end{table}
\end{frame}

\begin{frame}{Examples}
\protect\hypertarget{examples}{}
Model evolved by tournament selection after 100 generations:

@@

\begin{frame}{Research Question}
\protect\hypertarget{research-question-1}{}
\begin{itemize}
\tightlist
\item
  Does the usage of \(\epsilon\)-lexicase parent selection influence the
  generalization behaviour of genetic programming in symbolic regression
  if compared to tournament selection?
\end{itemize}

\begin{block}{Hypothesis testing}
\protect\hypertarget{hypothesis-testing}{}
\begin{block}{TODO: reformulate hypothesis}
\protect\hypertarget{todo-reformulate-hypothesis}{}
\begin{enumerate}
\tightlist
\item
  The usage of \(\epsilon\)-lexicase selection will result in models
  that perform significantly different than models that are evolved
  using tournament selection:
\item
  Statistical significant differences between the mean errors of
  training and testing data exist for both selection operators
\item
  No differences in program size exist between the distribution
  underlying the samples produced by \(\epsilon\)-lexicase and the
  distribution underlying samples of tournament selection
\end{enumerate}
\end{block}
\end{block}
\end{frame}

@@

\begin{frame}{Descriptive Statistics}
\protect\hypertarget{descriptive-statistics}{}
\begin{figure}
\centering
\includegraphics{../plots/mean_error_boxplot_all.png}
\caption{Distribution of Errors}
\end{figure}

@@

>>>>

\begin{table}

\caption{\label{tab:unnamed-chunk-1}Overview - Energy Heating data set}
\centering
\begin{tabular}[t]{l|l}
\hline
Variable & Description\\
\hline
X1 & Relative Compactness\\
\hline
X2 & Surface Area\\
\hline
X3 & Wall Area\\
\hline
X4 & Roof Area\\
\hline
X5 & Overall Height\\
\hline
X6 & Orientation\\
\hline
X7 & Glazing Area\\
\hline
X8 & Glazing Area Distribution\\
\hline
y1 & Heating Load\\
\hline
y2 & Cooling Load\\
\hline
\end{tabular}
\end{table}
\end{frame}

@@

\begin{frame}{Evolutionary Parameters}
\protect\hypertarget{evolutionary-parameters}{}
\begin{table}

\caption{\label{tab:unnamed-chunk-2}Evolutionary Parameters}
\centering
\begin{tabular}[t]{l|l}
\hline
Parameter & Value\\
\hline
Population Size & 500\\
\hline
Number of Generations & 100\\
\hline
Mutation Rate & 20\%\\
\hline
Crossover Rate & 80\%\\
\hline
Tournament Size & 3\\
\hline
Epsilon selection & automatic\\
\hline
Elite Size & 0\\
\hline
\end{tabular}
\end{table}
\end{frame}

\begin{frame}{Function Set}
\protect\hypertarget{function-set}{}
\begin{tabular}{l|r}
\hline
Function & Arity\\
\hline
Addition & 2\\
\hline
Subtraction & 2\\
\hline
Multiplication & 2\\
\hline
Negation & 1\\
\hline
Sine & 1\\
\hline
Cosine & 1\\
\hline
Protected Division & 2\\
\hline
Protected Natural Logarithm & 1\\
\hline
Protected Square Root & 1\\
\hline
\end{tabular}
\end{frame}

@@

\begin{frame}{Example}
\protect\hypertarget{example}{}
Model evolved by tournament selection after 100 generations:

@@

\begin{frame}{Research Question}
\protect\hypertarget{research-question-1}{}
\begin{itemize}
\tightlist
\item
  Does the usage of \(\epsilon\)-lexicase parent selection influence the
  generalization behavior of genetic programming in symbolic regression
  if compared to tournament selection?
\end{itemize}

\begin{block}{3 Questions}
\protect\hypertarget{questions}{}
\begin{enumerate}
\tightlist
\item
  Differences in average fitness for both algorithms?
\item
  Differences in fitness for training and testing data?
\item
  Differences in program size?
\end{enumerate}
\end{block}
\end{frame}

@@

\begin{frame}{Finding 1}
\protect\hypertarget{finding-1}{}
\begin{itemize}
\tightlist
\item
  The differences in average fitness of the final solutions between
  tournament selection and \(\epsilon\)-lexicase selection are highly
  statistical significant (\(\alpha=0.01\))
\item
  Tournament selection-based GP achieves a higher fitness for both
  training and testing data
\item
  Unexpected results based on the reviewed literature (La Cava, Spector
  and Danai, 2016), (La Cava \emph{et al.}, 2017)
\end{itemize}
\end{frame}

\begin{frame}{Statistical Test}
\protect\hypertarget{statistical-test}{}
\begin{table}

\caption{\label{tab:unnamed-chunk-5}Mean Error - P-Values (MWU)}
\centering
\begin{tabular}[t]{lrrrr}
\toprule
  & tournament\_training\_errors & tournament\_testing\_errors & elexicase\_training\_errors & elexicase\_testing\_errors\\
\midrule
tournament\_training\_errors & 1.000 & 0.309 & 0.000 & 0.000\\
tournament\_testing\_errors & 0.309 & 1.000 & 0.002 & 0.000\\
elexicase\_training\_errors & 0.000 & 0.002 & 1.000 & 0.257\\
elexicase\_testing\_errors & 0.000 & 0.000 & 0.257 & 1.000\\
\bottomrule
\end{tabular}
\end{table}
\end{frame}

\begin{frame}{Distribution of Fitness}
\protect\hypertarget{distribution-of-fitness}{}
\includegraphics{../plots/mean_error_boxplot_all.png}
\end{frame}

\begin{frame}{Finding 2}
\protect\hypertarget{finding-2}{}
\ldots{}
\end{frame}

\begin{frame}{Evolution of Fitness}
\protect\hypertarget{evolution-of-fitness}{}
\includegraphics{../plots/mean_error_combined.png}
\end{frame}

\begin{frame}{Evolution of Size}
\protect\hypertarget{evolution-of-size}{}
\begin{figure}
\centering
\includegraphics{../plots/size_subplotted.png}
\end{figure}

@@

\leavevmode\hypertarget{ref-https:ux2fux2fdoi.orgux2f10.48550ux2farxiv.1709.05394}{}%
La Cava, W. \emph{et al.} (2017) `A probabilistic and multi-objective
analysis of lexicase selection and epsilon-lexicase selection'. arXiv.
doi:\href{https://doi.org/10.48550/ARXIV.1709.05394}{10.48550/ARXIV.1709.05394}.
